All Equal in the Presence of Death? A Quantitative Analysis of the Early Iron Age Cemeteries of Knossos, Crete

### 2021-06-22 analysis/paper/Knossos_Cemeteries_Article_B&W.Rmd
Readme changes

#The appropriate packages are installed. Extrafont allows the use of non-standard fonts in R plots, ggplot2 is a powerful figure and graph plotting tool, and RColorBrewer includes a number of colour pallettes which can be applied to plots. The colour scheme in this paper is chosen from those that are suitable for colour-blind readers.
library(ggpubr)
library(ineq)
library(reshape2)
library(data.table)
library(ggplot2)
library(RColorBrewer)
library(grid)
library(dplyr)
library(plyr)
library(ggforce)
library(jpeg)
library(ggrepel)
library(forcats)
library(factoextra)
library(cluster)

@@

```{r tombconstruction, fig.dim=c(7,4), dpi=600, fig.cap="Tomb construction through time at both cemeteries. A) Counts based on the ceramic phase considered most likely to correspond to the tomb's first use (including simply 'PG', 'G' and 'O' where no greater specificity was possible). B) Aoristic sums of the same, where the uncertainty associated with general period assignations is spread across the corresponding sub-periods. In both cases, tombs lacking any securely dateable pottery are excluded."}

@@

new_tomb_counts_plot<-ggplot(data=subset(tombs,!is.na(start_period)),aes(start_period, fill=forcats::fct_rev(tomb_type_condensed),color=forcats::fct_rev(tomb_type_condensed)))+geom_bar(width=1,size=0.4)+
  scale_color_manual(values=c("black","black","black","black","black","black","black","black"),labels=c("Unknown","Shaft Grave","Pit Cave","Pithos Burial","Pit","Other","Larnax Burial","Chamber Tomb"))+
        theme_bw()+theme(legend.title=element_blank(),legend.spacing.x = unit(0.1,"cm"), text=element_text(family="serif", size=12),
          axis.text.x=element_text(angle=90, vjust=0.5, hjust=1),
          panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
            labs(x="Period", y="Count of new tombs")+scale_fill_manual(values=cbPalette,
                                                                                        labels=c("Unknown","Shaft Grave","Pit Cave",
                                                                                        "Pithos Burial","Pit","Other","Larnax Burial","Chamber Tomb"))+
             scale_y_continuous(expand=c(0,0),limits=c(0,25))+scale_x_discrete(expand=c(0,0))+
facet_wrap(~cem_condensed)+
               theme(strip.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA))

@@

The periods of construction shown in Fig. \@ref(fig:tombconstruction) correspond, therefore, largely with the earliest surviving pottery from each tomb. On this basis, it would appear that the KNC began with many tombs built during the - admittedly prolonged - SM period, followed by a protracted decline in construction, offset by resurgences in the LPG and LG periods. The Fortetsa tombs, which were all of chamber type, mostly date to PG, with almost no new tomb construction post-PGB. Indeed, it has been suggested that chamber tombs actually ceased to be created into the 8th century, with those appearing as MG or later in date being heavily looted and thus, in reality, older (Brock 1957, p.4; Cavanagh 1996, pp.652–53). What is clear is that many later tombs in the KNC are not chambers, but a mix of pit, *larnax*, and *pithos* burials, as well as unidentified forms. It is tempting to hypothesise similar burials having been overlooked at Fortetsa, but as a smaller cemetery it may have simply served a more restricted burying population and, thus, hosted more limited rites. The patterns at both cemeteries are by no means identical, then, even if there is some similarity in the declining investment in chamber tomb construction in the latter part of the EIA.

@@

```{r centsize,fig.dim=c(5,5), dpi=600, fig.cap="Scatterplots showing the relationship between the number of vessels dated to the A) 10th, B) 9th, C) 8th, and D) 7th centuries in tombs and the total number of vessels found therein. In each case, only tombs built by the century in question are plotted."}

@@

cbPalette <- c("#E69F00", "#56B4E9")

cor10th<-ggplot()+geom_point(data=subset(tombs,start_date>900),aes(vessels_10th,total_pottery,color=cem_condensed),size=0.95)+
  stat_cor(data=subset(tombs,start_date>900),aes(vessels_10th,total_pottery),method="pearson",
           label.x.npc=0.4,label.y.npc=0.9,family="serif",size=3,label.sep="\n")+
        theme_bw()+theme(text=element_text(family="serif", size=10),legend.title = element_blank(),
                       panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                       legend.text=element_text(size=10))+
              labs(y="Total vessels", x="10th century vessels")+
                               scale_x_continuous(expand = expansion(mult = c(0.001, 0.05)))+
                        scale_y_continuous(expand = expansion(mult = c(0.001, 0.05)),limits=c(0,470))+
  scale_color_manual(values=cbPalette)

cor9th<-ggplot()+geom_point(data=subset(tombs,start_date>800),aes(vessels_9th,total_pottery,color=cem_condensed),size=0.95)+
  stat_cor(data=subset(tombs,start_date>800),aes(vessels_9th,total_pottery),method="pearson",
           label.x.npc=0.7,label.y.npc=0.9,family="serif",size=3,label.sep="\n")+
        theme_bw()+theme(text=element_text(family="serif", size=10),legend.title = element_blank(),
                       panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                       legend.text=element_text(size=10))+
              labs(y="Total vessels", x="9th century vessels")+
                               scale_x_continuous(expand = expansion(mult = c(0.001, 0.05)))+
                        scale_y_continuous(expand = expansion(mult = c(0.001, 0.05)),limits=c(0,470))+
  scale_color_manual(values=cbPalette)

cor8th<-ggplot()+geom_point(data=subset(tombs,start_date>700),aes(vessels_8th,total_pottery,color=cem_condensed),size=0.95)+
  stat_cor(data=subset(tombs,start_date>700),aes(vessels_8th,total_pottery),method="pearson",
           label.x.npc=0.2,label.y.npc=0.9,family="serif",size=3,label.sep="\n")+
        theme_bw()+theme(text=element_text(family="serif", size=10),legend.title = element_blank(),
                       panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                       legend.text=element_text(size=10))+
              labs(y="Total vessels", x="8th century vessels")+
                               scale_x_continuous(expand = expansion(mult = c(0.001, 0.05)))+
                        scale_y_continuous(expand = expansion(mult = c(0.001, 0.05)),limits=c(0,470))+
  scale_color_manual(values=cbPalette)

cor7th<-ggplot()+geom_point(data=tombs,aes(vessels_7th,total_pottery,color=cem_condensed),size=0.95)+
  stat_cor(data=tombs,aes(vessels_7th,total_pottery),method="pearson",
           label.x.npc=0.6,label.y.npc=0.3,family="serif",size=3,label.sep="\n")+
        theme_bw()+theme(text=element_text(family="serif", size=10),legend.title = element_blank(),
                       panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                       legend.text=element_text(size=10))+
              labs(y="Total vessels", x="7th century vessels")+
                               scale_x_continuous(expand = expansion(mult = c(0.001, 0.05)))+
                        scale_y_continuous(expand = expansion(mult = c(0.001, 0.05)),limits=c(0,470))+
  scale_color_manual(values=cbPalette)

@@

This suspicion is well supported by Fig. \@ref(fig:centsize), where the strongest correlations are seen between tombs which received more 8th and 7th century vessels, and which had the largest assemblages overall. Therefore, the cemeteries' largest tombs seem a sensible starting point for investigating the dynamics of this late rise in burial activity.

@@

```{r cluster, fig.dim=c(3.5,4.1),dpi=600,fig.align='center',out.width="0.5%", fig.cap="Results of cluster analysis based on aoristic sums per decade for each tomb. Ward's method was used and tombs that yielded no pots at all were excluded. Four groups derived using a k-means clustering. Names/numbers of all tombs aside from those of Cluster 1 are shown." }

@@

fviz_cluster(list(data = df, cluster = sub_grp),ellipse.alpha=0.1,pointsize=1,
             show.clust.cent=FALSE,main=NULL,xlab="Dimension 1",ylab="Dimension 2",
             ggtheme=theme_bw()+theme(panel.grid.major = element_blank(),panel.grid.minor = element_blank(),
                                      text=element_text(family="serif", size=10),legend.position = "bottom"),
             geom="point",legend.text=element_text(size=10))+
  geom_text_repel(label=df_prep2$tomb,color="Black",family="serif",size=2.5)+
                scale_fill_manual('Cluster',values = cbPalette)+scale_color_manual('Cluster',values = cbPalette)+
  scale_shape_manual('Cluster', values=c(15,16,17,18))+scale_y_reverse()

@@

This analysis was done using Ward's method, and Fig. \@ref(fig:cluster) presents the results as a scatterplot, with the four groups separated by way of a k-means clustering. Cluster 1 comprises the vast majority of tombs in both cemeteries, while tombs Π (Pi) and 40 appear to be outliers. However, Clusters 3 and 4 reflect remarkably closely the preliminary observations above; Cluster 3 includes all those tombs which in Fig. \@ref(fig:toptombs) exhibit a late period of increased burial activity, while Cluster 4 largely comprises those with an earlier peak in depositions. All tombs in both clusters are chamber tombs, dating mostly to the 10th century in Cluster 4, and all to the 9th century in Cluster 3. Given that just the aoristic sums were submitted to analysis, it is encouraging that the results follow so closely an intuitive assessment of the data, and that tombs from both cemeteries appear in each group, suggesting from the outset that the differences between these may speak to social forces operating across the Knossian community. Tomb II is perhaps a notable absence, being the top-left-most point of Cluster 1, and we might choose to consider it a close relative of Cluster 3, but for the time being the results of the analysis will be adhered to. From now on, Clusters 3 and 4 shall be referred to as Groups A and B respectively.

@@

```{r tombgroupdimensions, fig.dim = c(7, 4),dpi=600, fig.align='center', fig.cap="A) Histograms showing the distribution of chamber tomb chamber areas, in square meters. B) The same but for dromos areas. In each case, these measurements are planimetric rather than volumetric, as accurate heights and depths for chambers and dromoi are so rarely recorded."}

chamber_area<-ggplot(data=subset(tombs, chamber_area>0),aes(chamber_area,fill=cluster,color=cluster))+
  geom_histogram(binwidth = 0.5,size=0.4)+
    theme_bw()+theme(text=element_text(family="serif", size=10),legend.position = "bottom",legend.title = element_blank(),
                     panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                     legend.text=element_text(size=10))+
            labs(x="Chamber area ("~m^2~")", y="Count of tombs per 0.5"~m^2~"")+
                        scale_fill_manual(values=cbPalette)+scale_color_manual(values=c("black","black","black"))+
                             scale_x_continuous(expand=expansion(mult=c(0,0.05)),breaks = seq(0, 10, by = 2))+
                  scale_y_continuous(expand=expansion(mult=c(0,0.05)),breaks = seq(0, 14, by = 2),limits=c(0,14))

dromos_area<-ggplot(data=subset(tombs, dromos_area>0),aes(dromos_area,fill=cluster,color=cluster))+
  geom_histogram(binwidth = 0.5,size=0.4)+
    theme_bw()+theme(text=element_text(family="serif", size=10),legend.position = "bottom",legend.title = element_blank(),
                     panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                     legend.text=element_text(size=10))+
            labs(x="Dromos area ("~m^2~")", y="Count of tombs per 0.5"~m^2~"")+
                        scale_fill_manual(values=cbPalette)+scale_color_manual(values=c("black","black","black"))+
                             scale_x_continuous(expand=expansion(mult=c(0,0.05)),breaks = seq(0, 12, by = 2))+
                  scale_y_continuous(expand=expansion(mult=c(0,0.05)),breaks = seq(0, 14, by = 2),limits=c(0,14))

@@

However, in considering those tombs where chamber and *dromos* area are calculable (Fig. \@ref(fig:tombgroupdimensions)), Groups A and B do not seem systematically marked out by their scale. The *dromos*, as the most visible area with regard to the funerary rite, might be considered the more indicative measure, and indeed three Group A and B tombs (75, 219 and 100) had very substantial *dromoi.* These features were also commonly disturbed by later activity, meaning other long *dromoi* may have gone unrecognised, but on the present evidence there is no general demarcation between the tombs based on size.

```{r tombdimensions, fig.dim = c(7, 4), dpi=600, fig.align='center', fig.cap="Scatterplots showing A) chamber areas of 10th century tombs, plotted against the number of 10th and 9th century vessels they yielded respectively. B) The same but using combined chamber and *dromos* areas. C) As A, but with tombs of the 9th century and vessels of the 9th and 8th centuries. D) As B, but with tombs of the 9th century and vessels of the 9th and 8th centuries."}

  cham_9th_pots_9th<-ggplot()+geom_point(data=subset(tombs,century_built=="9th"),aes(chamber_area,vessels_9th,color=cluster),size=1)+  stat_cor(data=subset(tombs,century_built=="9th"),aes(chamber_area,vessels_9th),method="pearson",label.x.npc=0.1,label.y.npc=0.9,family="serif",size=3,label.sep="\n")+
        theme_bw()+theme(text=element_text(family="serif", size=10),legend.position = "bottom",legend.title = element_blank(),
                       panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                       legend.text=element_text(size=10))+
              labs(x="Chamber area ("~m^2~")", y="9th century vessels")+
                          scale_color_manual(values=cbPalette)+
                               scale_x_continuous(expand = expansion(mult = c(0, 0.05)))+
                        scale_y_continuous(expand = expansion(mult = c(0, 0.05)))

  cham_9th_pots_8th<-ggplot()+geom_point(data=subset(tombs,century_built=="9th"),aes(chamber_area,vessels_8th,color=cluster),size=1)+  stat_cor(data=subset(tombs,century_built=="9th"),aes(chamber_area,vessels_8th),method="pearson",label.x.npc=0.6,label.y.npc=0.9,family="serif",size=3,label.sep="\n")+
        theme_bw()+theme(text=element_text(family="serif", size=10),legend.position = "bottom",legend.title = element_blank(),
                       panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                       legend.text=element_text(size=10))+
              labs(x="Chamber area ("~m^2~")", y="8th century vessels")+
                          scale_color_manual(values=cbPalette)+
                               scale_x_continuous(expand = expansion(mult = c(0, 0.05)))+
                        scale_y_continuous(expand = expansion(mult = c(0, 0.05)))

    dim_9th_pots_9th<-ggplot()+geom_point(data=subset(tombs,century_built=="9th"),aes(chamber_area+dromos_area,vessels_9th,color=cluster),size=1)+  stat_cor(data=subset(tombs,century_built=="9th"),aes(chamber_area+dromos_area,vessels_9th),method="pearson",label.x.npc=0.1,label.y.npc=0.9,family="serif",size=3,label.sep="\n")+
        theme_bw()+theme(text=element_text(family="serif", size=10),legend.position = "bottom",legend.title = element_blank(),
                       panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                       legend.text=element_text(size=10),axis.title.x = element_text(hjust=1))+
              labs(x="Dromos+chamber area ("~m^2~")", y="9th century vessels")+
                          scale_color_manual(values=cbPalette)+
                               scale_x_continuous(expand = expansion(mult = c(0, 0.05)))+
                        scale_y_continuous(expand = expansion(mult = c(0, 0.05)))

  dim_9th_pots_8th<-ggplot()+geom_point(data=subset(tombs,century_built=="9th"),aes(chamber_area+dromos_area,vessels_8th,color=cluster),size=1)+  stat_cor(data=subset(tombs,century_built=="9th"),aes(chamber_area+dromos_area,vessels_8th),method="pearson",label.x.npc=0.1,label.y.npc=0.8,family="serif",size=3,label.sep="\n")+
        theme_bw()+theme(text=element_text(family="serif", size=10),legend.position = "bottom",legend.title = element_blank(),
                       panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                       legend.text=element_text(size=10),axis.title.x = element_text(hjust=1))+
              labs(x="Dromos+chamber area ("~m^2~")", y="8th century vessels")+
                          scale_color_manual(values=cbPalette)+
                               scale_x_continuous(expand = expansion(mult = c(0, 0.05)))+
                        scale_y_continuous(expand = expansion(mult = c(0, 0.05)))

  cbPalette <- c("#009E73","#E69F00")

  cham_10th_pots_10th<-ggplot()+geom_point(data=subset(tombs,century_built=="10th"),aes(chamber_area,vessels_10th,color=cluster),size=1)+  stat_cor(data=subset(tombs,century_built=="10th"),aes(chamber_area,vessels_10th),method="pearson",label.x.npc=0.6,label.y.npc=0.9,family="serif",size=3,label.sep="\n")+
        theme_bw()+theme(text=element_text(family="serif", size=10),legend.position = "bottom",legend.title = element_blank(),
                       panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                       legend.text=element_text(size=10))+
              labs(x="Chamber area ("~m^2~")", y="10th century vessels")+
                          scale_color_manual(values=cbPalette)+
                               scale_x_continuous(expand = expansion(mult = c(0, 0.05)))+
                        scale_y_continuous(expand = expansion(mult = c(0, 0.05)))

>>>>

#The appropriate packages are installed. Extrafont allows the use of non-standard fonts in R plots, ggplot2 is a powerful figure and graph plotting tool, and RColorBrewer includes a number of colour pallettes which can be applied to plots. The colour scheme in this paper is chosen from those that are suitable for colour-blind readers.
library(ggpubr)
library(ineq)
library(reshape2)
library(data.table)
library(ggplot2)
library(RColorBrewer)
library(grid)
library(dplyr)
library(plyr)
library(ggforce)
library(jpeg)
library(ggrepel)
library(forcats)
library(factoextra)
library(cluster)
library(ggpattern)

@@

```{r tombconstructionbW, fig.dim=c(7,4), dpi=600, fig.cap="Tomb construction through time at both cemeteries. A) Counts based on the ceramic phase considered most likely to correspond to the tomb's first use (including simply 'PG', 'G' and 'O' where no greater specificity was possible). B) Aoristic sums of the same, where the uncertainty associated with general period assignations is spread across the corresponding sub-periods. In both cases, tombs lacking any securely dateable pottery are excluded."}

@@

new_tomb_counts_plot<-ggplot(data=subset(tombs,!is.na(start_period)),aes(start_period,pattern_type=forcats::fct_rev(tomb_type_condensed)))+geom_bar_pattern(pattern              = 'magick',
      pattern_fill         = 'black',
      pattern_aspect_ratio = 1.75,
      fill                 = 'white',
      colour               = 'black',width=1)+
        theme_bw()+theme(legend.title=element_blank(),legend.spacing.x = unit(0.1,"cm"), text=element_text(family="serif", size=12),
          axis.text.x=element_text(angle=90, vjust=0.5, hjust=1),
          panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
            labs(x="Period", y="Count of new tombs")+scale_pattern_type_discrete(choices = ggpattern::magick_pattern_names,labels=c("Unknown","Shaft Grave","Pit Cave","Pithos Burial","Pit","Other","Larnax Burial","Chamber Tomb"))+
             scale_y_continuous(expand=c(0,0),limits=c(0,25))+scale_x_discrete(expand=c(0,0))+
facet_wrap(~cem_condensed)+
               theme(strip.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA))

@@

The periods of construction shown in Fig. \@ref(fig:tombconstructionbw) correspond, therefore, largely with the earliest surviving pottery from each tomb. On this basis, it would appear that the KNC began with many tombs built during the - admittedly prolonged - SM period, followed by a protracted decline in construction, offset by resurgences in the LPG and LG periods. The Fortetsa tombs, which were all of chamber type, mostly date to PG, with almost no new tomb construction post-PGB. Indeed, it has been suggested that chamber tombs actually ceased to be created into the 8th century, with those appearing as MG or later in date being heavily looted and thus, in reality, older (Brock 1957, p.4; Cavanagh 1996, pp.652–53). What is clear is that many later tombs in the KNC are not chambers, but a mix of pit, *larnax*, and *pithos* burials, as well as unidentified forms. It is tempting to hypothesise similar burials having been overlooked at Fortetsa, but as a smaller cemetery it may have simply served a more restricted burying population and, thus, hosted more limited rites. The patterns at both cemeteries are by no means identical, then, even if there is some similarity in the declining investment in chamber tomb construction in the latter part of the EIA.

@@

```{r centsizebw,fig.dim=c(5,5), dpi=600, fig.cap="Scatterplots showing the relationship between the number of vessels dated to the A) 10th, B) 9th, C) 8th, and D) 7th centuries in tombs and the total number of vessels found therein. In each case, only tombs built by the century in question are plotted."}

@@

cor10th<-ggplot()+geom_point(data=subset(tombs,start_date>900),aes(vessels_10th,total_pottery,color=cem_condensed),size=0.95)+
  stat_cor(data=subset(tombs,start_date>900),aes(vessels_10th,total_pottery),method="pearson",
           label.x.npc=0.4,label.y.npc=0.9,family="serif",size=3,label.sep="\n")+
        theme_bw()+theme(text=element_text(family="serif", size=10),legend.title = element_blank(),
                       panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                       legend.text=element_text(size=10))+
              labs(y="Total vessels", x="10th century vessels")+
                               scale_x_continuous(expand = expansion(mult = c(0.001, 0.05)))+
                        scale_y_continuous(expand = expansion(mult = c(0.001, 0.05)),limits=c(0,470))+
  scale_color_grey(start=0.2,end=0.7)

cor9th<-ggplot()+geom_point(data=subset(tombs,start_date>800),aes(vessels_9th,total_pottery,color=cem_condensed),size=0.95)+
  stat_cor(data=subset(tombs,start_date>800),aes(vessels_9th,total_pottery),method="pearson",
           label.x.npc=0.7,label.y.npc=0.9,family="serif",size=3,label.sep="\n")+
        theme_bw()+theme(text=element_text(family="serif", size=10),legend.title = element_blank(),
                       panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                       legend.text=element_text(size=10))+
              labs(y="Total vessels", x="9th century vessels")+
                               scale_x_continuous(expand = expansion(mult = c(0.001, 0.05)))+
                        scale_y_continuous(expand = expansion(mult = c(0.001, 0.05)),limits=c(0,470))+
  scale_color_grey(start=0.2,end=0.7)

cor8th<-ggplot()+geom_point(data=subset(tombs,start_date>700),aes(vessels_8th,total_pottery,color=cem_condensed),size=0.95)+
  stat_cor(data=subset(tombs,start_date>700),aes(vessels_8th,total_pottery),method="pearson",
           label.x.npc=0.2,label.y.npc=0.9,family="serif",size=3,label.sep="\n")+
        theme_bw()+theme(text=element_text(family="serif", size=10),legend.title = element_blank(),
                       panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                       legend.text=element_text(size=10))+
              labs(y="Total vessels", x="8th century vessels")+
                               scale_x_continuous(expand = expansion(mult = c(0.001, 0.05)))+
                        scale_y_continuous(expand = expansion(mult = c(0.001, 0.05)),limits=c(0,470))+
  scale_color_grey(start=0.2,end=0.7)

cor7th<-ggplot()+geom_point(data=tombs,aes(vessels_7th,total_pottery,color=cem_condensed),size=0.95)+
  stat_cor(data=tombs,aes(vessels_7th,total_pottery),method="pearson",
           label.x.npc=0.6,label.y.npc=0.3,family="serif",size=3,label.sep="\n")+
        theme_bw()+theme(text=element_text(family="serif", size=10),legend.title = element_blank(),
                       panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                       legend.text=element_text(size=10))+
              labs(y="Total vessels", x="7th century vessels")+
                               scale_x_continuous(expand = expansion(mult = c(0.001, 0.05)))+
                        scale_y_continuous(expand = expansion(mult = c(0.001, 0.05)),limits=c(0,470))+
  scale_color_grey(start=0.2,end=0.7)

@@

This suspicion is well supported by Fig. \@ref(fig:centsizebw), where the strongest correlations are seen between tombs which received more 8th and 7th century vessels, and which had the largest assemblages overall. Therefore, the cemeteries' largest tombs seem a sensible starting point for investigating the dynamics of this late rise in burial activity.

@@

```{r clusterbw, fig.dim=c(3.5,4.1),dpi=600,fig.align='center',out.width="0.5%", fig.cap="Results of cluster analysis based on aoristic sums per decade for each tomb. Ward's method was used and tombs that yielded no pots at all were excluded. Four groups derived using a k-means clustering. Names/numbers of all tombs aside from those of Cluster 1 are shown." }

@@

fviz_cluster(list(data = df, cluster = sub_grp),ellipse.alpha=0.1,pointsize=1,
             show.clust.cent=FALSE,main=NULL,xlab="Dimension 1",ylab="Dimension 2",
             ggtheme=theme_bw()+theme(panel.grid.major = element_blank(),panel.grid.minor = element_blank(),
                                      text=element_text(family="serif", size=10),legend.position = "bottom"),
             geom="point",legend.text=element_text(size=10))+
  geom_text_repel(label=df_prep2$tomb,color="Black",family="serif",size=2.5)+
                scale_fill_grey('Cluster')+scale_color_grey('Cluster')+
  scale_shape_manual('Cluster', values=c(15,16,17,18))+scale_y_reverse()

@@

This analysis was done using Ward's method, and Fig. \@ref(fig:clusterbw) presents the results as a scatterplot, with the four groups separated by way of a k-means clustering. Cluster 1 comprises the vast majority of tombs in both cemeteries, while tombs Π (Pi) and 40 appear to be outliers. However, Clusters 3 and 4 reflect remarkably closely the preliminary observations above; Cluster 3 includes all those tombs which in Fig. \@ref(fig:toptombs) exhibit a late period of increased burial activity, while Cluster 4 largely comprises those with an earlier peak in depositions. All tombs in both clusters are chamber tombs, dating mostly to the 10th century in Cluster 4, and all to the 9th century in Cluster 3. Given that just the aoristic sums were submitted to analysis, it is encouraging that the results follow so closely an intuitive assessment of the data, and that tombs from both cemeteries appear in each group, suggesting from the outset that the differences between these may speak to social forces operating across the Knossian community. Tomb II is perhaps a notable absence, being the top-left-most point of Cluster 1, and we might choose to consider it a close relative of Cluster 3, but for the time being the results of the analysis will be adhered to. From now on, Clusters 3 and 4 shall be referred to as Groups A and B respectively.

@@

```{r tombgroupdimensionsbw, fig.dim = c(7, 4),dpi=600, fig.align='center', fig.cap="A) Histograms showing the distribution of chamber tomb chamber areas, in square meters. B) The same but for dromos areas. In each case, these measurements are planimetric rather than volumetric, as accurate heights and depths for chambers and dromoi are so rarely recorded."}

chamber_area<-ggplot(data=subset(tombs, chamber_area>0),aes(chamber_area,fill=cluster,color=cluster))+
  geom_histogram(binwidth = 0.5,size=0.4)+
    theme_bw()+theme(text=element_text(family="serif", size=10),legend.position = "bottom",legend.title = element_blank(),
                     panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                     legend.text=element_text(size=10))+
            labs(x="Chamber area ("~m^2~")", y="Count of tombs per 0.5"~m^2~"")+
                        scale_fill_grey(start=0.2,end=0.9)+scale_color_manual(values=c("black","black","black"))+
                             scale_x_continuous(expand=expansion(mult=c(0,0.05)),breaks = seq(0, 10, by = 2))+
                  scale_y_continuous(expand=expansion(mult=c(0,0.05)),breaks = seq(0, 14, by = 2),limits=c(0,14))

dromos_area<-ggplot(data=subset(tombs, dromos_area>0),aes(dromos_area,fill=cluster,color=cluster))+
  geom_histogram(binwidth = 0.5,size=0.4)+
    theme_bw()+theme(text=element_text(family="serif", size=10),legend.position = "bottom",legend.title = element_blank(),
                     panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                     legend.text=element_text(size=10))+
            labs(x="Dromos area ("~m^2~")", y="Count of tombs per 0.5"~m^2~"")+
                        scale_fill_grey(start=0.2,end=0.9)+scale_color_manual(values=c("black","black","black"))+
                             scale_x_continuous(expand=expansion(mult=c(0,0.05)),breaks = seq(0, 12, by = 2))+
                  scale_y_continuous(expand=expansion(mult=c(0,0.05)),breaks = seq(0, 14, by = 2),limits=c(0,14))

@@

However, in considering those tombs where chamber and *dromos* area are calculable (Fig. \@ref(fig:tombgroupdimensionsbw)), Groups A and B do not seem systematically marked out by their scale. The *dromos*, as the most visible area with regard to the funerary rite, might be considered the more indicative measure, and indeed three Group A and B tombs (75, 219 and 100) had very substantial *dromoi.* These features were also commonly disturbed by later activity, meaning other long *dromoi* may have gone unrecognised, but on the present evidence there is no general demarcation between the tombs based on size.

```{r tombdimensionsbw, fig.dim = c(7, 4), dpi=600, fig.align='center', fig.cap="Scatterplots showing A) chamber areas of 10th century tombs, plotted against the number of 10th and 9th century vessels they yielded respectively. B) The same but using combined chamber and *dromos* areas. C) As A, but with tombs of the 9th century and vessels of the 9th and 8th centuries. D) As B, but with tombs of the 9th century and vessels of the 9th and 8th centuries."}

  cham_9th_pots_9th<-ggplot()+geom_point(data=subset(tombs,century_built=="9th"),aes(chamber_area,vessels_9th,color=cluster),size=1)+  stat_cor(data=subset(tombs,century_built=="9th"),aes(chamber_area,vessels_9th),method="pearson",label.x.npc=0.1,label.y.npc=0.9,family="serif",size=3,label.sep="\n")+
        theme_bw()+theme(text=element_text(family="serif", size=10),legend.position = "bottom",legend.title = element_blank(),
                       panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                       legend.text=element_text(size=10))+
              labs(x="Chamber area ("~m^2~")", y="9th century vessels")+
                          scale_color_grey(start=0.2,end=0.8)+
                               scale_x_continuous(expand = expansion(mult = c(0, 0.05)))+
                        scale_y_continuous(expand = expansion(mult = c(0, 0.05)))

  cham_9th_pots_8th<-ggplot()+geom_point(data=subset(tombs,century_built=="9th"),aes(chamber_area,vessels_8th,color=cluster),size=1)+  stat_cor(data=subset(tombs,century_built=="9th"),aes(chamber_area,vessels_8th),method="pearson",label.x.npc=0.6,label.y.npc=0.9,family="serif",size=3,label.sep="\n")+
        theme_bw()+theme(text=element_text(family="serif", size=10),legend.position = "bottom",legend.title = element_blank(),
                       panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                       legend.text=element_text(size=10))+
              labs(x="Chamber area ("~m^2~")", y="8th century vessels")+
                          scale_color_grey(start=0.2,end=0.8)+
                               scale_x_continuous(expand = expansion(mult = c(0, 0.05)))+
                        scale_y_continuous(expand = expansion(mult = c(0, 0.05)))

    dim_9th_pots_9th<-ggplot()+geom_point(data=subset(tombs,century_built=="9th"),aes(chamber_area+dromos_area,vessels_9th,color=cluster),size=1)+  stat_cor(data=subset(tombs,century_built=="9th"),aes(chamber_area+dromos_area,vessels_9th),method="pearson",label.x.npc=0.1,label.y.npc=0.9,family="serif",size=3,label.sep="\n")+
        theme_bw()+theme(text=element_text(family="serif", size=10),legend.position = "bottom",legend.title = element_blank(),
                       panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                       legend.text=element_text(size=10),axis.title.x = element_text(hjust=1))+
              labs(x="Dromos+chamber area ("~m^2~")", y="9th century vessels")+
                          scale_color_grey(start=0.2,end=0.8)+
                               scale_x_continuous(expand = expansion(mult = c(0, 0.05)))+
                        scale_y_continuous(expand = expansion(mult = c(0, 0.05)))

  dim_9th_pots_8th<-ggplot()+geom_point(data=subset(tombs,century_built=="9th"),aes(chamber_area+dromos_area,vessels_8th,color=cluster),size=1)+  stat_cor(data=subset(tombs,century_built=="9th"),aes(chamber_area+dromos_area,vessels_8th),method="pearson",label.x.npc=0.1,label.y.npc=0.8,family="serif",size=3,label.sep="\n")+
        theme_bw()+theme(text=element_text(family="serif", size=10),legend.position = "bottom",legend.title = element_blank(),
                       panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                       legend.text=element_text(size=10),axis.title.x = element_text(hjust=1))+
              labs(x="Dromos+chamber area ("~m^2~")", y="8th century vessels")+
                          scale_color_grey(start=0.2,end=0.8)+
                               scale_x_continuous(expand = expansion(mult = c(0, 0.05)))+
                        scale_y_continuous(expand = expansion(mult = c(0, 0.05)))

  cham_10th_pots_10th<-ggplot()+geom_point(data=subset(tombs,century_built=="10th"),aes(chamber_area,vessels_10th,color=cluster),size=1)+  stat_cor(data=subset(tombs,century_built=="10th"),aes(chamber_area,vessels_10th),method="pearson",label.x.npc=0.6,label.y.npc=0.9,family="serif",size=3,label.sep="\n")+
        theme_bw()+theme(text=element_text(family="serif", size=10),legend.position = "bottom",legend.title = element_blank(),
                       panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                       legend.text=element_text(size=10))+
              labs(x="Chamber area ("~m^2~")", y="10th century vessels")+
                          scale_color_grey(start=0.5,end=0.8)+
                               scale_x_continuous(expand = expansion(mult = c(0, 0.05)))+
                        scale_y_continuous(expand = expansion(mult = c(0, 0.05)))

  cham_10th_pots_9th<-ggplot()+geom_point(data=subset(tombs,century_built=="10th"),aes(chamber_area,vessels_9th,color=cluster),size=1)+  stat_cor(data=subset(tombs,century_built=="10th"),aes(chamber_area,vessels_9th),method="pearson",label.x.npc=0.1,label.y.npc=0.85,family="serif",size=3,label.sep="\n")+
        theme_bw()+theme(text=element_text(family="serif", size=10),legend.position = "bottom",legend.title = element_blank(),
                       panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                       legend.text=element_text(size=10))+
              labs(x="Chamber area ("~m^2~")", y="9th century vessels")+
                          scale_color_grey(start=0.5,end=0.8)+
                               scale_x_continuous(expand = expansion(mult = c(0, 0.05)))+
                        scale_y_continuous(expand = expansion(mult = c(0, 0.05)))
